\plaintitle

### 2020-03-26 wip-design/paper.tex
References, case studies

@@

>>>>

Evaluating ui systems research
\cite{evaluating}

Type providers
\cite{inforich,fsdata,dotdriven}

Assistants for SW / auto-complete
\cite{assistants}

Journalism tools
\cite{idyll}

Notebook systems
\cite{jupyter} or more transparency \cite{wrattler}
managing messes \cite{messes}

Visual data exploration tools
\cite{control,tableau,vizdom}

Wrangler
\cite{wrangler}

Victor
\cite{learnable,principle}

Live and exploratory programming
\cite{review} - review of the two
\cite{exploratory}, which led to new tools~\cite{variolite} (supports informal versioning),

\cite{liveroad}
review the use of live programming in a Smalltalk derived environment. Lighttable~\cite{lighttable}

A more principled approach can be used by systems based on structured editing~\cite{structure-based,livenut,lamdu}

Subtext
\cite{subtext,directprog}

sketch and sketch
\cite{sketchnsketch}

lower this barrier to valid, replicablestatistical analysis, we introduce Tea1, a high-level declar-ative language and runtime system
(still JSON, but good aims)
\cite{tea}

DS.js: Turn Any Webpage into an Example-Centric Live Programming Environment for Learning Data Science
(cool but just JS)
\cite{dsjs}

Eviza: A Natural Language Interface for Visual Analysis
(from Tableau; uses auto-complete to suggest natural language phrases)
\cite{eviza}

Declarative Interaction Design for Data Visualization
(interactive construction of vega-like grammar of graphics)
\cite{interactionviz}

programming by demonstration
\cite{watch}

"Dontcheva et al. [4] use PBD techniques to enable
extraction, integration, and templated search of web data.
Gulwani [8] introduces an algorithm for learning expres-
sive string processing programs from input-output examples"

programming by example as an alternative
\cite{spreadsheetpbe,flashextract}

Lyra: An interactive visualization design environment
\cite{lyra}

end-user programming
\cite{enduser} - overview, \cite{smallmatter} book

touch based data science??? must inevitably have fewer intraections
NA

autocomplete / predictive text work
\cite{predictive} - predicting data transformation scripts in a somewhat more complex language
\cite{proactive} -

\cite{potter}
Potter’s Wheel: An Interactive Data Cleaning System
n interactive data clean-ing system that tightly integrates transformation anddiscrepancy detection.   Users  gradually build  trans-formations  to  clean  the  data  by  adding  or  undoingtransforms on a spreadsheet-like interface; the effectof a transform is shown at once on records visible onscreen.

\newpage

@@

The Gamma aims to simplify programmatic data exploration while keeping enough expressive power
to allow users to create interesting data visualizations. We consider the expressive power in
this section and assess simplicity in the next one. We used The Gamma to analyse
the UK government expenditure, activities of a research institute, Olympic medals and
information about the Doctor Who series\footnote{Links removed to preserve anonymity.}.
In this section, we consider two interesting larger examples.

\subsection{If Michael Phelps were a Country}
Michael Phelps has won so many medals that media compared him to
countries.\footnote{\href{https://www.npr.org/sections/thetorch/2016/08/14/489832779/}{\small\bf\ttfamily npr.org/sections/thetorch/2016/08/14/489832779/}}
Using the tabular data provider, we construct a chart shown in Figure~\ref{fig:cases} that
produces a country league table including Michael Phelps:\footnote{Available at: \url{http://gallery.thegamma.net/86/}}

\begin{thegamma}
\kvd{let} data = olympics
    .'group data'.'by Team'.'sum Gold'.then
    .'sort data'.'by Gold descending'.then
    .paging.skip(43).take(4)
    .'get series'.'with key Team'.'and value Gold'

\kvd{let} phelps = olympics
  .'filter data'.'Athlete is'.'Michael Phelps'.then
  .'group data'.'by Athlete'.'sum Gold'.then
  .'get series'.'with key Athlete'.'and value Gold'

charts.bar(data.append(phelps))
  .setColors(["#aec7e8","#aec7e8","#1f77b4"])
\end{thegamma}

The core of the data analysis is done in two commands. The first counts gold medals by countries
and uses paging to fetch 4 countries with suitable number of medals. In the second, we abuse the
grouping operation to aggregate data for just a single group. The two data series are then assigned
to local variables (for readability) and passed to the \ikvd{chart.columns} function.
In this case study, the core data exploration can be constructed via iterative prompting, but
building a chart requires invoking the \ikvd{append} operation and specifying colors using a list.

\subsection{The Most Frequent Doctor Who Villains}
In the second case study, we list Doctor Who villains by the number of episodes in which they
appear.\footnote{Available at: \url{http://gallery.thegamma.net/87/}} This is interesting as it
combines the graph database provider for fetching the data with the tabular data provider for
summarization:

\begin{thegamma}
\kvd{let} topEnemies = drWho.Character.Doctor
  .'ENEMY OF'.'[any]'.'APPEARED IN'.'[any]'.explore
  .'group data'.'by Character name'
    .'count distinct Episode name'.then
  .'sort data'.'by Episode name descending'.then
  .paging.take(8).'get series'
    .'with key Character name'.'and value Episode name'

chart.bar(topEnemies)
\end{thegamma}

Lines 1-2 use the graph database provider to find all paths linking the Doctor node with any
character linked via \ikvd{ENEMY OF}, followed by any episode linked by \ikvd{APPEARED IN}.
This produces a table that can be analysed using the tabular data provider by choosing the
\ikvd{explore} member. For each character (the villain) we count the total number of
distinct episodes in the table. The result is shown in Figure~\ref{fig:cases}.

The logic is not trivial, but it performs a fairly sophisticated data analysis that involves a
graph database query followed by an SQL-like data aggregation. The code can be constructed
using iterative prompting (with the exception of the numbers in paging). It can also be constructed
gradually and instantaneous prviews support the user while doing so.

\newpage

\begin{figure}
\centering
\includegraphics[width=1\columnwidth]{figures/cases}
\caption{Charts produced by two The Gamma case studies.}
\label{fig:cases}
\end{figure}

~
\newpage
